---
title: "Basis Functions for the Baseline Hazard"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Basis Functions for the Baseline Hazard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(quiet = TRUE)
}
```

## The MPL baseline hazard expansion

The Cox proportional hazards model specifies the hazard for subject $i$ as

$$h(t_i) = h_0(t_i)\,e^{\mathbf{x}_i^T \boldsymbol{\beta}},$$

where $\boldsymbol{\beta}$ collects the regression coefficients and $h_0(\cdot)$ is
an unspecified baseline hazard.  Maximum penalised likelihood (MPL) estimation
approximates $h_0$ by a finite mixture of non-negative *basis functions*
$\psi_1, \ldots, \psi_m$:

$$h_0(t) = \sum_{u=1}^{m} \theta_u\,\psi_u(t), \qquad
H_0(t) = \int_{t_{(1)}}^{t} h_0(v)\,dv = \sum_{u=1}^{m} \theta_u\,\Psi_u(t),$$

where $\Psi_u(t) = \int_{t_{(1)}}^{t}\psi_u(v)\,dv$ is the cumulative basis
and $\boldsymbol{\theta} = (\theta_1,\ldots,\theta_m)^T \geq \mathbf{0}$ are
coefficients estimated alongside $\boldsymbol{\beta}$ [@Ma2014;@MaWebbHudson2024].

`survivalMPL` provides four production basis types, selectable via the `basis`
argument of `coxph_mpl.control()`:

| `basis` string | Description |
|---|---|
| `"uniform"` (or `"u"`) | Piecewise-constant (step function) |
| `"gaussian"` (or `"g"`) | Truncated Gaussian kernels |
| `"msplines"` (or `"m"`) | M-splines of order $o$ |
| `"epanechikov"` (or `"e"`, `"epanechnikov"`) | Epanechnikov kernels |

The rest of this article presents the mathematical definition of each basis,
visualises the shapes, and compares fits on the bundled `melanoma` dataset.

---

## Uniform (step-function) basis

The time range $[t_{(1)}, t_{(n)}]$ is partitioned into $m$ adjacent intervals
$[\alpha_u, \alpha_{u+1})$.  The density and cumulative bases are

$$\psi_u(t) = \mathbf{1}(\alpha_u \leq t < \alpha_{u+1}),$$

$$\Psi_u(t) = \mathbf{1}(\alpha_u \leq t)
  \Bigl[t^{\mathbf{1}(t < \alpha_{u+1})}\,\alpha_{u+1}^{\mathbf{1}(\alpha_{u+1} \leq t)}
        - \alpha_u\Bigr].$$

Each $\theta_u$ is then interpreted directly as the constant hazard rate on
$[\alpha_u, \alpha_{u+1})$.

---

## Gaussian basis

Basis function $u$ is a truncated Gaussian density centred at knot $\alpha_u$
with scale $\sigma_u$:

$$\psi_u(t) = \frac{1}{\sigma_u\,\delta_u}\,
  \phi\!\left(\frac{t - \alpha_u}{\sigma_u}\right), \qquad
\Psi_u(t) = \frac{1}{\delta_u}
  \left[\Phi\!\left(\frac{t - \alpha_u}{\sigma_u}\right)
       -\Phi\!\left(\frac{t_{(1)} - \alpha_u}{\sigma_u}\right)\right],$$

where $\phi$ and $\Phi$ are the standard normal density and CDF, and
$\delta_u = \Phi\!\left(\tfrac{t_{(n)}-\alpha_u}{\sigma_u}\right)
           -\Phi\!\left(\tfrac{t_{(1)}-\alpha_u}{\sigma_u}\right)$
ensures each truncated basis integrates to 1 on $[t_{(1)}, t_{(n)}]$.

---

## M-spline basis

M-splines of order $o$ are built recursively from augmented knots
$\boldsymbol{\alpha}^\star = \bigl[\alpha_1 \mathbf{1}_{o-1}^T,\,\boldsymbol{\alpha}^T,\,\alpha_{n_\alpha}\mathbf{1}_{o-1}^T\bigr]^T$:

$$\psi^o_u(t) =
\begin{cases}
\dfrac{\mathbf{1}(\alpha^\star_u \leq t < \alpha^\star_{u+1})}
      {\alpha^\star_{u+1} - \alpha^\star_u} & o = 1,\\[8pt]
\dfrac{o}{o-1}
\dfrac{\mathbf{1}(\alpha^\star_u \leq t < \alpha^\star_{u+o})}
      {\alpha^\star_{u+o} - \alpha^\star_u}
\Bigl[(t-\alpha^\star_u)\,\psi^{o-1}_u(t)
     +(\alpha^\star_{u+o}-t)\,\psi^{o-1}_{u+1}(t)\Bigr] & \text{otherwise.}
\end{cases}$$

The cumulative $\Psi^o_u(t)$ is an I-spline (integrated M-spline); each
$\psi^o_u$ integrates to 1 over its support $[\alpha^\star_u, \alpha^\star_{u+o}]$.
`survivalMPL` defaults to $o = 3$ (cubic M-splines), for which the second-order
roughness penalty has a closed form.

---

## Epanechnikov basis

Epanechnikov basis functions have compact parabolic support.  For the interior
functions ($1 < u < m$):

$$\psi_u(t) = \frac{6\,\mathbf{1}(\alpha^\star_u \leq t < \alpha^\star_{u+o})
  (t - \alpha^\star_u)(t - \alpha^\star_{u+o})}{(\alpha^\star_u - \alpha^\star_{u+o})^3}.$$

The endpoint functions ($u = 1$ and $u = m$) use asymmetric variants to ensure a
proper probability density on their support.  Like M-splines, these use the
augmented knot sequence $\boldsymbol{\alpha}^\star$.

---

## Visual comparison

The code below uses the package's internal basis machinery to evaluate
$\psi_u(t)$ on a fine grid and plot each set of basis functions.

```{r basis-vis, fig.height=9, out.width="100%", echo=FALSE, message=FALSE}
library(survivalMPL)
library(survival)

# ---- event times from the melanoma dataset for knot placement ----
data(melanoma)
events <- sort(c(melanoma$t_L[melanoma$t_L > 0],
                 melanoma$t_R[is.finite(melanoma$t_R)]))
t_grid <- seq(min(events), max(events), length.out = 400)

n_obs <- sum(melanoma$t_L == melanoma$t_R & is.finite(melanoma$t_R))
ctrl_u <- coxph_mpl.control(basis = "uniform",     n.obs = n_obs)
ctrl_g <- coxph_mpl.control(basis = "gaussian",    n.obs = n_obs)
ctrl_m <- coxph_mpl.control(basis = "msplines",    n.obs = n_obs)
ctrl_e <- coxph_mpl.control(basis = "epanechikov", n.obs = n_obs)

knots_u <- survivalMPL:::compute_knots(ctrl_u, events)
knots_g <- survivalMPL:::compute_knots(ctrl_g, events)
knots_m <- survivalMPL:::compute_knots(ctrl_m, events)
knots_e <- survivalMPL:::compute_knots(ctrl_e, events)

# Crop x to the 95th-percentile of event times — most basis support lies here.
# The melanoma data is right-skewed (90% of events fall below ~0.9 years).
t_vis_max <- quantile(events, 1)
t_vis     <- seq(min(events[events > 0]), t_vis_max, length.out = 400)

psi_u <- survivalMPL:::compute_basis_matrix(t_vis, knots_u, "uniform",     ctrl_u$order, which = 1)
psi_g <- survivalMPL:::compute_basis_matrix(t_vis, knots_g, "gaussian",    ctrl_g$order, which = 1)
psi_m <- survivalMPL:::compute_basis_matrix(t_vis, knots_m, "msplines",    ctrl_m$order, which = 1)
psi_e <- survivalMPL:::compute_basis_matrix(t_vis, knots_e, "epanechikov", ctrl_e$order, which = 1)

# ---- plot ----
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# The bases produce 11-23 functions from the default knots, which is far too
# many to read in one panel.  Draw an evenly spaced subset of `n_show` of them
# instead: enough to see the shape and the overlap, still spanning the whole
# time range.  The legend gives the indices actually plotted.
plot_basis <- function(mat, t, title, knots_alpha, n_show = 6) {
  idx   <- unique(round(seq(1, ncol(mat), length.out = min(n_show, ncol(mat)))))
  mat   <- mat[, idx, drop = FALSE]
  cols  <- seq_along(idx)
  # With only a handful of columns the old 0.99 cap is set by the boundary
  # functions, which are several times taller than the rest; 0.95 keeps the
  # interior bumps filling the panel.
  y_cap <- quantile(mat[mat > 0 & is.finite(mat)], 0.95, na.rm = TRUE) * 1.1
  matplot(t, mat, type = "l", lty = 1, col = cols, lwd = 1.6,
          xlab = "t (years)", ylab = expression(psi[u](t)),
          main = title, ylim = c(0, y_cap))
  abline(v = knots_alpha[knots_alpha <= max(t)], col = "grey70", lty = 2)
  legend("topright",
         legend = paste0("ψ", idx),
         col = cols, lty = 1, lwd = 1.6,
         bty = "n", cex = 0.75)
}

plot_basis(psi_u, t_vis, "Uniform",      knots_u$Alpha)
plot_basis(psi_g, t_vis, "Gaussian",     knots_g$Alpha)
plot_basis(psi_m, t_vis, "M-Splines",    knots_m$Alpha)
plot_basis(psi_e, t_vis, "Epanechnikov", knots_e$Alpha)
```

Dashed vertical lines mark the internal knots $\boldsymbol{\alpha}$.

---

## Fitting comparison on `melanoma`

The choice of basis affects the smoothness and shape of the estimated baseline
hazard but rarely changes regression coefficient estimates substantially.
Here we fit the same interval-censored model with all four bases using the
bundled pseudo-melanoma data (time to first local recurrence, $n = 300$, true
baseline $h_0(t) = t^{-1/2}$).

```{r mel-prep}
data(melanoma)
formula_mel <- Surv(t_L, t_R, type = "interval2") ~
  Arm + Leg + Trunk + mm1to2 + mm2to4 + mm4plus + Female + Age_centred
n_obs <- sum(melanoma$t_L == melanoma$t_R & is.finite(melanoma$t_R))
base_ctrl <- list(n.obs = n_obs,
                  smooth = 0,
                  max.iter = c(40, 2e4, 5e4))
```

```{r mel-fit-all}
fit_u <- coxph_mpl(formula_mel, data = melanoma,
  control = do.call(coxph_mpl.control,
    c(base_ctrl, list(basis = "uniform"))))

fit_g <- coxph_mpl(formula_mel, data = melanoma,
  control = do.call(coxph_mpl.control,
    c(base_ctrl, list(basis = "gaussian"))))

fit_m <- coxph_mpl(formula_mel, data = melanoma,
  control = do.call(coxph_mpl.control,
    c(base_ctrl, list(basis = "msplines"))))

fit_e <- coxph_mpl(formula_mel, data = melanoma,
  control = do.call(coxph_mpl.control,
    c(base_ctrl, list(basis = "epanechikov"))))
```

### Regression coefficients

```{r coef-table}
coef_mat <- cbind(
  Uniform      = coef(fit_u),
  Gaussian     = coef(fit_g),
  `M-Splines`  = coef(fit_m),
  Epanechnikov = coef(fit_e)
)
round(coef_mat, 3)
```

### Estimated baseline hazards

```{r baseline-plot, fig.height=4.5, out.width="100%", echo=FALSE}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

fits   <- list(fit_u, fit_g, fit_m, fit_e)
labels <- c("Uniform", "Gaussian", "M-Splines", "Epanechnikov")
cols   <- c("steelblue", "tomato", "forestgreen", "darkorchid")

t_min <- quantile(events, 0.02)
t_seq <- seq(t_min, max(events), length.out = 300)

# instantaneous risk at mean covariates
risks <- sapply(fits, function(f) predict(f, type = "risk", time = t_seq)$risk)
y_max <- quantile(risks, 0.995) * 1.1
plot(NA, xlim = range(t_seq), ylim = c(0, y_max),
     xlab = "Time (years)", ylab = expression(hat(h)(t)),
     main = "Instantaneous risk (mean covariates)", cex.main = 0.9)
for (i in seq_along(fits))
  lines(t_seq, risks[, i], col = cols[i], lwd = 1.5)
legend("topright",
       legend = labels,
       col = cols, lty = 1,
       lwd = 1.5, bty = "n", cex = 0.75)

# survival at mean covariates
plot(NA, xlim = range(t_seq), ylim = c(0, 1),
     xlab = "Time (years)", ylab = expression(hat(S)(t)),
     main = "Survival (mean covariates)", cex.main = 0.9)
for (i in seq_along(fits)) {
  p <- predict(fits[[i]], type = "survival", time = t_seq)
  lines(p$time, p$survival, col = cols[i], lwd = 1.5)
}
legend("topright",
       legend = labels,
       col = cols, lty = 1,
       lwd = 1.5, bty = "n", cex = 0.75)
```

All four bases agree on the broad shape — a hazard that is high early and decays
over time — and on the implied survival curve.  The smooth bases (Gaussian,
M-Splines, Epanechnikov) produce continuous estimates; the uniform basis
introduces visible step artefacts that diminish with more knots.

Note that these are predictions at the *mean* covariate vector, so each curve
estimates $h_0(t)\exp(\bar{\mathbf{x}}^T\boldsymbol{\beta})$ rather than $h_0(t)$
itself.  The covariates are not centred, so this is not directly comparable to
the Weibull baseline used to simulate the data.

---

## Practical guidance

- **`"msplines"` (default for interval-censored data)** - M-splines offer
  a good balance of smoothness and computational efficiency.  The second-order
  roughness penalty has an exact closed form.

- **`"uniform"`** - fast and simple; useful as a baseline comparison.  Use more
  knots (`n.knots`) if the hazard is not close to piecewise-constant.

- **`"gaussian"`** - smooth, kernel-based; both first- and second-order
  roughness penalties are available.  Can be slow with many knots.

- **`"epanechnikov"`** - compact support like M-splines but with parabolic
  rather than polynomial shape; only the second-order penalty is supported.

---

## Extensibility

The basis registry (`R/basis.R`) allows new basis types to be added without
modifying existing code.  The file `R/basis-bsplines.R` includes a skeleton 
showing how to implement the three required functions (`knots_fn`, `matrix_fn`, 
`penalty_fn`) and how to call `register_basis()`.  
See `?list_bases` for the full list of registered bases.

---

## References
