---
title: "Control Parameters"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Control Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5
)
set.seed(1)
```

## Overview

`coxph_mpl()` runs with defaults that work for most datasets, so the tutorials
call it with nothing but a formula and a data frame.  When a fit needs tuning -
a different baseline hazard basis, more or fewer knots, a fixed smoothing
value, longer iteration limits - the settings live in
`coxph_mpl.control()`.

There are two equivalent ways to change them:

```{r two-ways, eval=FALSE}
# 1. pass control arguments directly; coxph_mpl() forwards them
coxph_mpl(Surv(time, status) ~ x, data = df, basis = "msplines", tol = 1e-5)

# 2. build the control object explicitly
coxph_mpl(Surv(time, status) ~ x, data = df,
          control = coxph_mpl.control(n.obs = sum(df$status),
                                      basis = "msplines", tol = 1e-5))
```

The first form is shorter and is what the other articles use.  The second is
needed when the same settings are reused across several fits.  Note the
`n.obs` argument in the second form: `coxph_mpl.control()` uses the number of
fully observed events to set defaults, and cannot count them itself, so a
hand-built control object must be told.  Called through `...`, `coxph_mpl()`
counts them for you.

A control object is a plain list with validated entries:

```{r control-object}
library(survivalMPL)
library(survival)

str(coxph_mpl.control(n.obs = 165))
```

Every argument is validated on the way in: values outside their admissible
range are silently replaced by the default rather than passed on to the
optimiser, so an impossible setting cannot produce a meaningless fit.

---

## The baseline hazard

### `basis`

Which family of non-negative functions $\psi_1,\ldots,\psi_m$ is used to build
$h_0(t) = \sum_u \theta_u \psi_u(t)$:

| Value | Aliases | Shape |
|---|---|---|
| `"uniform"` | `"u"`, `"uni"` | Piecewise-constant step functions (default) |
| `"msplines"` | `"m"`, `"ms"` | M-splines of order `order` |
| `"gaussian"` | `"g"`, `"gauss"` | Truncated Gaussian kernels |
| `"epanechikov"` | `"e"`, `"epa"`, `"epanechnikov"` | Epanechnikov kernels |

`"uniform"` is the fastest and gives coefficients $\theta_u$ that read directly
as the hazard rate on knot interval $u$.  `"msplines"` is the usual choice when
a smooth $\hat h_0$ matters, in particular for interval-censored data where
$H_0$ enters the likelihood at both endpoints.  **Basis Functions for the
Baseline Hazard** defines each one and compares them on the same data.

### `n.knots` and `range.quant`

`n.knots` is a length-2 vector `c(k1, k2)`.  The first entry places $k_1 + 1$
knots at quantiles of the observed event times up to `range.quant[2]` (the 90th
percentile by default); the second places $k_2 + 1$ equally spaced knots above
that quantile.  When `k2 = 0`, all knots are quantile-based over the whole
range.  Defaults are `c(8, 2)` for `"uniform"` and `"msplines"`, and `c(0, 20)`
for the two kernel bases.

More knots mean a more flexible $\hat h_0$ and a larger $\boldsymbol{\theta}$
to estimate; the penalty, not the knot count, is what should control
smoothness, so a moderately generous knot count with automatic smoothing is usually
better than a small one:

```{r knots, cache=FALSE}
lung <- na.omit(lung[, c("time", "status", "age", "sex", "ph.karno", "wt.loss")])
f    <- Surv(time, status == 2) ~ age + sex + ph.karno + wt.loss

fit_coarse <- coxph_mpl(f, data = lung, n.knots = c(3, 1),  tol = 1e-5)
fit_fine   <- coxph_mpl(f, data = lung, n.knots = c(15, 3), tol = 1e-5)

c(coarse = fit_coarse$knots$m, fine = fit_fine$knots$m)
round(cbind(coarse = coef(fit_coarse), fine = coef(fit_fine)), 4)
```

The regression coefficients barely move; what changes is the resolution of the
baseline hazard.

```{r knots-plot, fig.width=8, fig.height=4}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(fit_coarse, which = 2, ask = FALSE, cex.main = 0.85)
plot(fit_fine,   which = 2, ask = FALSE, cex.main = 0.85)
```

### `order` and `penalty`

`order` is the order of the M-spline or Epanechnikov basis (default 3, cubic
M-splines).  Order 1 M-splines coincide with the uniform basis and order 2 with
a triangular basis.

`penalty` is the order of the roughness penalty: 1 penalises the first
differences of $\boldsymbol{\theta}$, 2 the second differences.  The uniform and
Gaussian bases accept either; the Epanechnikov basis always uses second order,
and M-splines use `order - 1`.  The default is 2.

---

## Smoothing: `smooth`

The penalised log-likelihood maximised by `coxph_mpl()` is

$$\Phi(\boldsymbol{\beta},\boldsymbol{\theta}) =
  \ell(\boldsymbol{\beta},\boldsymbol{\theta}) -
  \lambda\,\boldsymbol{\theta}^T \mathbf{R} \boldsymbol{\theta},$$

where $\mathbf{R}$ is the roughness matrix of the chosen basis and $\lambda \ge 0$
is the smoothing value [@Ma2014].  `smooth` sets $\lambda$:

| `smooth` | Effect |
|---|---|
| `NULL` (default) | $\lambda$ selected automatically from the data |
| `0` | No penalty at all: plain maximum likelihood |
| any $\lambda > 0$ | $\lambda$ held fixed at that value |

**Automatic selection** is the default, and is the marginal likelihood method of
@MaWebbHudson2024 (Section 2.10).  The quadratic penalty
$\boldsymbol{\theta}^T\mathbf{R}\boldsymbol{\theta}$ is read as a normal prior
$\boldsymbol{\theta} \sim N(\mathbf{0}, \sigma^2\mathbf{R}^{-1})$ with
$\sigma^2 = 1/(2\lambda)$; $\boldsymbol\beta$ and $\boldsymbol\theta$ are
integrated out, and a Laplace approximation to the resulting marginal
likelihood gives the update
$\hat\sigma^2 = \hat{\boldsymbol\theta}^T\mathbf{R}\hat{\boldsymbol\theta}/\nu$,
where $\nu$ is the model degrees of freedom.  Since the estimates and $\sigma^2$
depend on each other, this alternates: **inner iterations** fit
$\boldsymbol\beta$ and $\boldsymbol\theta$ at the current $\sigma^2$, then an
**outer iteration** updates $\sigma^2$, until $\nu$ stabilises.

Whether $\lambda$ was selected or fixed is visible in the fitted object:
supplying `smooth` sets `max.iter[1]` to 1, which switches the outer iterations
off, and `summary()` reports the value as *fixed* rather than *estimated*.  The
final $\lambda$ - selected or not - is stored in `fit$control$smooth`.

```{r smooth-fits}
fit_ml   <- coxph_mpl(f, data = lung, smooth = 0)      # unpenalised
fit_auto <- coxph_mpl(f, data = lung)                  # automatic smoothing value
fit_hard <- coxph_mpl(f, data = lung, smooth = 1e8)    # heavily penalised

c(ml = fit_ml$control$smooth,
  automatic = fit_auto$control$smooth,
  hard = fit_hard$control$smooth)
```

### Choosing $\lambda$ by hand

$\lambda$ is not scale-free.  It trades off against the size of
$\boldsymbol{\theta}$, which depends on the time units, the basis and the number
of knots, so the value that oversmooths one dataset is negligible on another -
the selected value above is of order $10^7$ only because `lung` measures time in
days, making the hazard rates themselves of order $10^{-3}$.  There is no
portable rule of thumb, which is the reason to leave `smooth = NULL` unless
there is a specific reason not to.  If a fixed value is needed, fit a small
grid and inspect $\hat h_0$, rather than transplanting a value from another
analysis.  Extremely large values eventually break the fit rather than
flattening it: on `lung`, $\lambda = 10^9$ drives the estimate to a degenerate
solution with an attenuated `sex` coefficient.

---

## The optimiser

| Argument | Meaning | Default |
|---|---|---|
| `max.iter` | Length-3 vector: (1) outer smoothing-parameter updates, (2) inner $\boldsymbol\beta$/$\boldsymbol\theta$ updates, (3) total inner iterations | `c(150, 7.5e4, 1e6)` |
| `tol` | Convergence tolerance: on the parameter change between inner iterations, and, as `10 * tol`, on the change in the degrees of freedom $\nu$ between outer iterations | `1e-7` |
| `kappa` | Step-size reduction factor (> 1) applied when the penalised likelihood fails to increase | `1/0.6` |
| `min.theta` | $\hat\theta_u$ below this are reported as zero, i.e. as active non-negativity constraints | `1e-10` |
| `epsilon` | Length-2 floor on survival and baseline hazard values, guarding the logarithms | `c(1e-16, 1e-10)` |

Setting `smooth` to a fixed value forces `max.iter[1]` to 1, since there is no
outer loop left to run.  `summary()` prints whether the algorithm converged and
how many outer and inner iterations it used; `fit$iter` holds the same two
numbers:

```{r converged}
fit_auto$iter        # c(outer iterations, total inner iterations)
```

If either equals its cap the fit has not converged, and `summary()` says
`Convergence: NO`.  `fit_auto` above is such a fit: with everything at its
default it stops at the cap of 150 outer iterations without the degrees of
freedom having settled.

The reason is the stopping rule rather than the data.  The outer loop stops when
$\nu$ changes by less than `10 * tol`, which is $10^{-6}$ at the default
`tol = 1e-7`, whereas @MaWebbHudson2024 describe stability in $\nu$ as changes
"not greater than, say, 1 or 0.5".  Asking for six decimal places of a quantity
the method itself treats as settled at half a unit is what keeps the loop
running.  Either loosening `tol` to the $10^{-5}$ the book uses in its own
examples, or raising the cap, produces the same fit:

```{r converged-more}
fit_tol  <- coxph_mpl(f, data = lung, tol = 1e-05)
fit_more <- coxph_mpl(f, data = lung, max.iter = c(600, 7.5e4, 1e6))

rbind(`tol = 1e-05`  = fit_tol$iter,
      `max.iter 600` = fit_more$iter)

round(cbind(`default (capped)` = coef(fit_auto),
            `tol = 1e-05`      = coef(fit_tol),
            `max.iter 600`     = coef(fit_more)), 4)
```

The three sets of coefficients differ by at most a twentieth of a standard
error, so the capped fit was not wrong - but a fit that reports non-convergence
should not be published on the strength of that, and `tol = 1e-05` is the
cheaper of the two ways out.  If neither helps, the usual causes are too many knots for the number
of events, or a covariate on a very different scale from the others.

---

## Ties: `ties` and `seed`

Knot sequences are built from the distinct fully observed event times.
Duplicated event times therefore have to be resolved:

- `ties = "epsilon"` (default) jitters duplicates by a small random amount;
- `ties = "unique"` drops duplicates before computing quantiles.

Because `"epsilon"` uses the random number generator, `seed` fixes it so the
knots - and hence the fit - are exactly reproducible:

```{r ties}
fit_a <- coxph_mpl(f, data = lung, ties = "epsilon", seed = 42)
fit_b <- coxph_mpl(f, data = lung, ties = "epsilon", seed = 42)
identical(coef(fit_a), coef(fit_b))
```

The current RNG state is restored afterwards, so setting `seed` does not
disturb any simulation running around the fit.

---

## Full argument list

| Argument | Purpose |
|---|---|
| `n.obs` | Number of fully observed events; used to derive defaults |
| `basis` | Basis for $h_0$ |
| `smooth` | Smoothing value $\lambda$; `NULL` = automatic selection, `0` = no penalty |
| `max.iter` | Iteration limits for the three optimisation stages |
| `tol` | Convergence tolerance |
| `n.knots` | Quantile and equally spaced internal knot counts |
| `range.quant` | Quantile range for the quantile-based knots |
| `order` | Order of the M-spline / Epanechnikov basis |
| `penalty` | Order of the roughness penalty |
| `cover.sigma.quant`, `cover.sigma.fixed` | Bandwidth targets for the Gaussian basis |
| `min.theta` | Threshold below which $\hat\theta_u$ is reported as zero |
| `kappa` | Step-size reduction factor |
| `epsilon` | Numerical floors for survival and hazard values |
| `ties` | Tie-breaking strategy for duplicated event times |
| `seed` | RNG seed used when `ties = "epsilon"` |
| `n.events_basis` | Events per uniform basis element; retained for backward compatibility, not used by the current knot construction |

See `?coxph_mpl.control` for the full argument documentation.

---

## Practical recipes

- **Start with the defaults.**  Uniform basis, automatic smoothing, default knots.
- **Interval- or left-censored data:** `basis = "msplines"`, since $H_0$ is
  evaluated at both interval endpoints and a step function makes that estimate
  coarse.
- **Comparing bases or reproducing a maximum likelihood fit:** `smooth = 0`.
- **Left truncation (`entry`):** `basis = "uniform"` - the only basis for which
  differencing the cumulative basis is implemented.  Any other basis is
  replaced by `"uniform"` with a warning.
- **A fit that will not converge:** set `tol = 1e-05` first - the outer loop
  stopping rule at the default tolerance is stricter than the method needs -
  then raise `max.iter`, then reduce `n.knots`.

## References
