---
title: "Comparing coxph() and coxph_mpl()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Comparing coxph() and coxph_mpl()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5
)
set.seed(1)
```

## Overview

`survival::coxph()` maximises the *partial likelihood* - a function of
regression coefficients $\boldsymbol{\beta}$ only, with the baseline hazard
$h_0(t)$ profiled out.  `survivalMPL::coxph_mpl()` maximises the *penalised
full likelihood*, estimating $\boldsymbol{\beta}$ and $h_0(t)$ jointly via a
smooth basis expansion.

For right-censored data with non-informative censoring the two approaches
produce essentially the same $\hat{\boldsymbol{\beta}}$ and standard errors
[@Ma2014], while `coxph_mpl` additionally delivers a smooth, closed-form
$\hat{h}_0(t)$ that can be plotted and predicted without additional
post-processing.

This tutorial uses the `lung` dataset (`survival` package) - 228 patients with
advanced lung cancer, right-censored overall survival to compare:

1. Regression coefficient estimates and standard errors.
2. 95 % confidence intervals (forest plot).
3. Estimated baseline survival: Kaplan-Meier, Breslow (from `coxph`), and
   M-spline (from `coxph_mpl`).

---

## Data and model fits

```{r fits}
library(survivalMPL)
library(survival)

lung <- na.omit(lung[, c("time", "status", "age", "sex", "ph.karno", "wt.loss")])
f    <- Surv(time, status == 2) ~ age + sex + ph.karno + wt.loss

# --- partial likelihood (coxph) ---
fit_pl <- coxph(f, data = lung, x = TRUE)

# --- penalised full likelihood, M-splines (coxph_mpl) ---
fit_mpl <- coxph_mpl(f, data = lung,
  control = coxph_mpl.control(
    basis    = "msplines",
    n.obs    = sum(lung$status == 2),
    max.iter = c(150, 7.5e4, 1e6)
  )
)
```

---

## Regression coefficients and standard errors

```{r coef-table}
beta_pl  <- coef(fit_pl)
se_pl    <- sqrt(diag(vcov(fit_pl)))

beta_mpl <- coef(fit_mpl)
se_mpl   <- fit_mpl$se$Beta$M2HM2

row_labels <- c(
  age      = "Age (per year)^a^",
  sex      = "Sex",
  ph.karno = "Karnofsky score (per point)^a^",
  wt.loss  = "Weight loss (kg)"
)

tab <- data.frame(
  Covariate    = row_labels[names(beta_pl)],
  `PL est.`    = round(beta_pl,  4),
  `PL SE`      = round(se_pl,    4),
  `MPL est.`   = round(beta_mpl, 4),
  `MPL SE`     = round(se_mpl,   4),
  `Difference` = round(beta_mpl - beta_pl, 4),
  check.names  = FALSE,
  row.names    = NULL
)
knitr::kable(tab, align = "lrrrrr",
             caption = "^a^ The forest plot rescales these covariates to per-10-unit changes.")
```

The estimates agree closely: the full-likelihood approach recovers the same
regression signal as partial likelihood, with near-identical standard errors.
The `MPL SE` reported here uses the sandwich variance estimator for a fixed
smoothing parameter derived in @Ma2014 (equation 29), accessed via
`se = "M2HM2"` in the package.

---

## Forest plot

```{r forest-plot, echo=FALSE, fig.height=5, fig.width=10, out.width="100%"}
# Scale age and ph.karno to per-10-unit changes for comparable magnitudes
scale  <- c(age = 10, sex = 1, ph.karno = 10, wt.loss = 1)
labels <- c(age = "Age (per 10 yrs)", sex = "Sex (F vs M)",
            ph.karno = "Karnofsky score\n(per 10 pts)", wt.loss = "Weight loss (kg)")

b_pl  <- beta_pl  * scale
b_mpl <- beta_mpl * scale
s_pl  <- se_pl    * scale
s_mpl <- se_mpl   * scale

z      <- qnorm(0.975)
lo_pl  <- b_pl  - z * s_pl;  hi_pl  <- b_pl  + z * s_pl
lo_mpl <- b_mpl - z * s_mpl; hi_mpl <- b_mpl + z * s_mpl

p     <- length(b_pl)
y_pl  <- seq(p, 1) + 0.18
y_mpl <- seq(p, 1) - 0.18

# right-side annotation text: "β [lo, hi]"
fmt <- function(b, lo, hi)
  sprintf("%.3f  [%.3f, %.3f]", b, lo, hi)
ann_pl  <- mapply(fmt, b_pl,  lo_pl,  hi_pl)
ann_mpl <- mapply(fmt, b_mpl, lo_mpl, hi_mpl)

cols <- c(pl = "steelblue", mpl = "tomato")

# layout: narrow left margin, wide right margin for annotations
par(mar = c(4, 9, 4, 1))
x_lim <- range(c(lo_pl, hi_pl, lo_mpl, hi_mpl))
x_lim <- x_lim + c(-1, 1) * diff(x_lim) * 0.05

plot(NA,
     xlim = x_lim, ylim = c(0.3, p + 0.7),
     xlab = "Coefficient (log-HR, scaled)", ylab = "",
     yaxt = "n", main = "Regression coefficients ± 1.96 SE")

abline(v = 0, col = "grey50", lty = 2)
abline(h = seq_len(p), col = "grey92", lty = 1)   # row guides

axis(2, at = seq(p, 1), labels = rev(labels), las = 1, cex.axis = 0.88)

# PL
segments(lo_pl, y_pl, hi_pl, y_pl, col = cols["pl"], lwd = 2)
points(b_pl, y_pl, pch = 16, col = cols["pl"], cex = 1.2)

# MPL
segments(lo_mpl, y_mpl, hi_mpl, y_mpl, col = cols["mpl"], lwd = 2)
points(b_mpl, y_mpl, pch = 17, col = cols["mpl"], cex = 1.2)

# header for annotations
mtext("β  [95% CI]", side = 3, adj = 1, cex = 0.75, col = "grey30", line = 0.3)

# right-side numeric annotations (outside the plot via xpd)
x_ann <- par("usr")[2] + diff(par("usr")[1:2]) * 0.02
for (i in seq_len(p)) {
  ri <- p + 1 - i   # reverse: top row = last covariate
  mtext(ann_pl[i],  side = 4, at = y_pl[ri],  las = 1,
        col = cols["pl"],  cex = 0.65, line = 0.3)
  mtext(ann_mpl[i], side = 4, at = y_mpl[ri], las = 1,
        col = cols["mpl"], cex = 0.65, line = 0.3)
}

legend("bottomleft",
       legend = c("coxph (PL)", "coxph_mpl (MPL, M2HM2)"),
       col = cols, pch = c(16, 17), lty = 1, lwd = 2,
       bty = "n", cex = 0.85)
```

---

## Baseline survival comparison

Three survival estimators are overlaid:

- **Kaplan-Meier** - nonparametric marginal estimate, no covariate adjustment.
  Serves as a visual reference for the raw data.
- **Breslow** - step-function baseline from `coxph` evaluated at mean
  covariates via `survfit()`.  This is the standard Cox output for
  $\hat{S}(t \mid \bar{\mathbf{x}})$.
- **M-splines (MPL)** - smooth closed-form estimate from `coxph_mpl`,
  also evaluated at mean covariates.

Breslow and M-splines estimate the same quantity 
$\hat{S}(t \mid \bar{\mathbf{x}})$ and should track each other closely.
Kaplan-Meier sits apart because it does not adjust for covariates.

```{r baseline-surv, echo=FALSE, fig.height=5, out.width="90%"}
# Kaplan-Meier (marginal, unadjusted)
km <- survfit(Surv(time, status == 2) ~ 1, data = lung)

# Breslow baseline at mean covariates
mean_nd <- as.data.frame(lapply(lung[, c("age","sex","ph.karno","wt.loss")], mean))
sf_pl   <- survfit(fit_pl, newdata = mean_nd)

# MPL M-splines at mean covariates (i = NULL defaults to column means of X)
t_seq    <- seq(min(lung$time), max(lung$time), length.out = 400)
pred_mpl <- predict(fit_mpl, type = "survival", time = t_seq)

par(mar = c(4, 4, 3, 1))
plot(km,
     col = "grey60", lwd = 1.5, conf.int = FALSE,
     xlab = "Time (days)", ylab = expression(hat(S)(t)),
     main = "Survival: KM (marginal) vs Breslow vs M-splines (mean covariates)",
     xlim = c(0, max(lung$time)), ylim = c(0, 1))

lines(sf_pl$time, sf_pl$surv, col = "steelblue", lwd = 2)
lines(pred_mpl$time, pred_mpl$survival, col = "tomato", lwd = 2)

legend("topright",
       legend = c("Kaplan-Meier (marginal, unadjusted)",
                  "Breslow / coxph (mean covariates)",
                  "M-splines / coxph_mpl (mean covariates)"),
       col    = c("grey60", "steelblue", "tomato"),
       lwd    = c(1.5, 2, 2),
       bty    = "n", cex = 0.85)
```

The Breslow estimate is a step function that jumps only at observed event times.
The M-spline estimate is smooth by construction and closely tracks the Breslow
curve, while avoiding the staircase artefacts that arise from tied or sparse
event times.  The Kaplan-Meier curve sits above both because it is not adjusted
for covariates - in particular it does not account for the protective effect of
female sex and higher Karnofsky score at the mean covariate profile.

---

## Instantaneous hazard comparison

Survival curves are the forgiving comparison: integration smooths away
differences in the hazard.  The instantaneous hazard $\hat h(t)$ is where the
two approaches genuinely differ, and it is the quantity `coxph()` does not
estimate at all.

The partial likelihood gives a cumulative hazard that jumps at each event time,
so its "hazard" is a set of point masses.  To draw it as a rate it has to be
binned: the increments of the Breslow cumulative hazard are accumulated over
equal-width intervals and divided by the interval width, which is the
piecewise-constant hazard estimate implied by `coxph()` at that resolution.
`coxph_mpl()` needs no such post-processing - $\hat h(t)$ is part of the fitted
object.

```{r hazard-comparison, echo=FALSE, fig.height=5, out.width="90%"}
# --- binned Breslow hazard at mean covariates ---
H_pl  <- -log(sf_pl$surv)                       # cumulative hazard, step function
brks  <- seq(0, max(lung$time), length.out = 9) # 8 equal-width bins
H_brk <- approx(c(0, sf_pl$time), c(0, H_pl), xout = brks, rule = 2)$y
h_bin <- diff(H_brk) / diff(brks)

# --- smooth MPL hazard at mean covariates ---
t_haz    <- seq(min(lung$time), max(lung$time), length.out = 400)
haz_mpl  <- predict(fit_mpl, type = "risk", time = t_haz)

par(mar = c(4, 4.2, 3, 1))
plot(NA, xlim = c(0, max(lung$time)), ylim = c(0, max(c(h_bin, haz_mpl$risk)) * 1.15),
     xlab = "Time (days)", ylab = expression(hat(h)(t)),
     main = "Instantaneous hazard: binned Breslow vs M-splines (mean covariates)",
     cex.main = 0.95, las = 1)
segments(brks[-length(brks)], h_bin, brks[-1], h_bin, col = "steelblue", lwd = 2.4)
segments(brks[-c(1, length(brks))], h_bin[-length(h_bin)],
         brks[-c(1, length(brks))], h_bin[-1], col = "steelblue", lwd = 1, lty = 3)
lines(haz_mpl$time, haz_mpl$risk, col = "tomato", lwd = 2.4)
rug(lung$time[lung$status == 2], col = "grey60")

legend("topleft",
       legend = c("Breslow / coxph, binned into 8 intervals",
                  "M-splines / coxph_mpl (smooth)",
                  "observed death times"),
       col = c("steelblue", "tomato", "grey60"),
       lwd = c(2.4, 2.4, 1), lty = c(1, 1, 1), bty = "n", cex = 0.85)
```

Over the bulk of follow-up the two agree: a hazard of roughly $1.3 \times
10^{-3}$ per day early on, rising to around $3 \times 10^{-3}$ by day 500.  The
disagreement is in the character of the two estimates rather than their level.

The binned Breslow estimate jumps from bin to bin - its fourth bin sits below
its third - and the whole picture depends on a bin width nobody chose on
statistical grounds: eight bins give this, sixteen would give something visibly
rougher, four something visibly flatter.  Its last bin is exactly zero, because
no death is recorded after day 894, which is a statement about this sample
rather than about the hazard.  The M-spline estimate is smooth by construction,
with the roughness penalty rather than an arbitrary bin width deciding how much
structure survives, and every point of it carries a standard error - which the
binned version does not.

Past about 900 days the two part company completely: the smooth estimate goes
on rising while the binned one collapses to zero.  Neither is informative
there.  Three patients are still at risk and none of them dies, so the data
contain almost nothing about the hazard at that point, and each method fills
the vacuum in its own way - one by extrapolating its trend, the other by
reporting the empty bin as a zero rate.

That difference is the practical argument for the penalised full likelihood:
when the target is a hazard ratio, `coxph()` and `coxph_mpl()` agree, and the
partial likelihood is the cheaper route; when the target is absolute risk - a
hazard, a survival probability, a predicted event time for a given covariate
profile - the baseline has to be estimated, and `coxph_mpl()` estimates it as
part of the model.

---

## References
