---
title: "Comparing coxph and coxph_mpl"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Comparing coxph and coxph_mpl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`survival::coxph()` maximises the *partial likelihood* - a function of
regression coefficients $\boldsymbol{\beta}$ only, with the baseline hazard
$h_0(t)$ profiled out.  `survivalMPL::coxph_mpl()` maximises the *penalised
full likelihood*, estimating $\boldsymbol{\beta}$ and $h_0(t)$ jointly via a
smooth basis expansion.

For right-censored data with non-informative censoring the two approaches
produce essentially the same $\hat{\boldsymbol{\beta}}$ and standard errors
[@Ma2014], while `coxph_mpl` additionally delivers a smooth, closed-form
$\hat{h}_0(t)$ that can be plotted and predicted without additional
post-processing.

This tutorial uses the `lung` dataset (`survival` package) - 228 patients with
advanced lung cancer, right-censored overall survival to compare:

1. Regression coefficient estimates and standard errors.
2. 95 % confidence intervals (forest plot).
3. Estimated baseline survival: Kaplan–Meier, Breslow (from `coxph`), and
   M-spline (from `coxph_mpl`).

---

## Data and model fits

```{r fits}
library(survivalMPL)
library(survival)

lung <- na.omit(lung[, c("time", "status", "age", "sex", "ph.karno", "wt.loss")])
f    <- Surv(time, status == 2) ~ age + sex + ph.karno + wt.loss

# --- partial likelihood (coxph) ---
fit_pl <- coxph(f, data = lung, x = TRUE)

# --- penalised full likelihood, M-splines (coxph_mpl) ---
fit_mpl <- coxph_mpl(f, data = lung,
  control = coxph_mpl.control(
    basis    = "msplines",
    n.obs    = sum(lung$status == 2),
    max.iter = c(150, 7.5e4, 1e6)
  )
)
```

---

## Regression coefficients and standard errors

```{r coef-table}
beta_pl  <- coef(fit_pl)
se_pl    <- sqrt(diag(vcov(fit_pl)))

beta_mpl <- coef(fit_mpl)
se_mpl   <- fit_mpl$se$Beta$M2HM2

row_labels <- c(
  age      = "Age (per year)^a^",
  sex      = "Sex",
  ph.karno = "Karnofsky score (per point)^a^",
  wt.loss  = "Weight loss (kg)"
)

tab <- data.frame(
  Covariate   = row_labels[names(beta_pl)],
  `PL β`      = round(beta_pl,  4),
  `PL SE`     = round(se_pl,    4),
  `MPL β`     = round(beta_mpl, 4),
  `MPL SE`    = round(se_mpl,   4),
  `Δβ`        = round(beta_mpl - beta_pl, 4),
  check.names = FALSE,
  row.names   = NULL
)
knitr::kable(tab, align = "lrrrrr",
             caption = "^a^ The forest plot rescales these covariates to per-10-unit changes.")
```

The estimates agree closely: the full-likelihood approach recovers the same
regression signal as partial likelihood, with near-identical standard errors.
The `MPL SE` reported here uses the sandwich variance estimator for a fixed
smoothing parameter derived in @Ma2014 (equation 29), accessed via
`se = "M2HM2"` in the package.

---

## Forest plot

```{r forest-plot, echo=FALSE, fig.height=5, fig.width=10, out.width="100%"}
# Scale age and ph.karno to per-10-unit changes for comparable magnitudes
scale  <- c(age = 10, sex = 1, ph.karno = 10, wt.loss = 1)
labels <- c(age = "Age (per 10 yrs)", sex = "Sex (F vs M)",
            ph.karno = "Karnofsky score\n(per 10 pts)", wt.loss = "Weight loss (kg)")

b_pl  <- beta_pl  * scale
b_mpl <- beta_mpl * scale
s_pl  <- se_pl    * scale
s_mpl <- se_mpl   * scale

z      <- qnorm(0.975)
lo_pl  <- b_pl  - z * s_pl;  hi_pl  <- b_pl  + z * s_pl
lo_mpl <- b_mpl - z * s_mpl; hi_mpl <- b_mpl + z * s_mpl

p     <- length(b_pl)
y_pl  <- seq(p, 1) + 0.18
y_mpl <- seq(p, 1) - 0.18

# right-side annotation text: "β [lo, hi]"
fmt <- function(b, lo, hi)
  sprintf("%.3f  [%.3f, %.3f]", b, lo, hi)
ann_pl  <- mapply(fmt, b_pl,  lo_pl,  hi_pl)
ann_mpl <- mapply(fmt, b_mpl, lo_mpl, hi_mpl)

cols <- c(pl = "steelblue", mpl = "tomato")

# layout: narrow left margin, wide right margin for annotations
par(mar = c(4, 9, 4, 1))
x_lim <- range(c(lo_pl, hi_pl, lo_mpl, hi_mpl))
x_lim <- x_lim + c(-1, 1) * diff(x_lim) * 0.05

plot(NA,
     xlim = x_lim, ylim = c(0.3, p + 0.7),
     xlab = "Coefficient (log-HR, scaled)", ylab = "",
     yaxt = "n", main = "Regression coefficients ± 1.96 SE")

abline(v = 0, col = "grey50", lty = 2)
abline(h = seq_len(p), col = "grey92", lty = 1)   # row guides

axis(2, at = seq(p, 1), labels = rev(labels), las = 1, cex.axis = 0.88)

# PL
segments(lo_pl, y_pl, hi_pl, y_pl, col = cols["pl"], lwd = 2)
points(b_pl, y_pl, pch = 16, col = cols["pl"], cex = 1.2)

# MPL
segments(lo_mpl, y_mpl, hi_mpl, y_mpl, col = cols["mpl"], lwd = 2)
points(b_mpl, y_mpl, pch = 17, col = cols["mpl"], cex = 1.2)

# header for annotations
mtext("β  [95% CI]", side = 3, adj = 1, cex = 0.75, col = "grey30", line = 0.3)

# right-side numeric annotations (outside the plot via xpd)
x_ann <- par("usr")[2] + diff(par("usr")[1:2]) * 0.02
for (i in seq_len(p)) {
  ri <- p + 1 - i   # reverse: top row = last covariate
  mtext(ann_pl[i],  side = 4, at = y_pl[ri],  las = 1,
        col = cols["pl"],  cex = 0.65, line = 0.3)
  mtext(ann_mpl[i], side = 4, at = y_mpl[ri], las = 1,
        col = cols["mpl"], cex = 0.65, line = 0.3)
}

legend("bottomleft",
       legend = c("coxph (PL)", "coxph_mpl (MPL, M2HM2)"),
       col = cols, pch = c(16, 17), lty = 1, lwd = 2,
       bty = "n", cex = 0.85)
```

---

## Baseline survival comparison

Three survival estimators are overlaid:

- **Kaplan–Meier** - nonparametric marginal estimate, no covariate adjustment.
  Serves as a visual reference for the raw data.
- **Breslow** - step-function baseline from `coxph` evaluated at mean
  covariates via `survfit()`.  This is the standard Cox output for
  $\hat{S}(t \mid \bar{\mathbf{x}})$.
- **M-splines (MPL)** - smooth closed-form estimate from `coxph_mpl`,
  also evaluated at mean covariates.

Breslow and M-splines estimate the same quantity 
$\hat{S}(t \mid \bar{\mathbf{x}})$ and should track each other closely.
Kaplan–Meier sits apart because it does not adjust for covariates.

```{r baseline-surv, echo=FALSE, fig.height=5, out.width="90%"}
# Kaplan-Meier (marginal, unadjusted)
km <- survfit(Surv(time, status == 2) ~ 1, data = lung)

# Breslow baseline at mean covariates
mean_nd <- as.data.frame(lapply(lung[, c("age","sex","ph.karno","wt.loss")], mean))
sf_pl   <- survfit(fit_pl, newdata = mean_nd)

# MPL M-splines at mean covariates (i = NULL defaults to column means of X)
t_seq    <- seq(min(lung$time), max(lung$time), length.out = 400)
pred_mpl <- predict(fit_mpl, type = "survival", time = t_seq)

par(mar = c(4, 4, 3, 1))
plot(km,
     col = "grey60", lwd = 1.5, conf.int = FALSE,
     xlab = "Time (days)", ylab = expression(hat(S)(t)),
     main = "Survival: KM (marginal) vs Breslow vs M-splines (mean covariates)",
     xlim = c(0, max(lung$time)), ylim = c(0, 1))

lines(sf_pl$time, sf_pl$surv, col = "steelblue", lwd = 2)
lines(pred_mpl$time, pred_mpl$survival, col = "tomato", lwd = 2)

legend("topright",
       legend = c("Kaplan-Meier (marginal, unadjusted)",
                  "Breslow / coxph (mean covariates)",
                  "M-splines / coxph_mpl (mean covariates)"),
       col    = c("grey60", "steelblue", "tomato"),
       lwd    = c(1.5, 2, 2),
       bty    = "n", cex = 0.85)
```

The Breslow estimate is a step function that jumps only at observed event times.
The M-spline estimate is smooth by construction and closely tracks the Breslow
curve, while avoiding the staircase artefacts that arise from tied or sparse
event times.  The Kaplan–Meier curve sits above both because it is not adjusted
for covariates - in particular it does not account for the protective effect of
female sex and higher Karnofsky score at the mean covariate profile.

---

## References
