---
title: "Getting Started with survivalMPL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with survivalMPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`survivalMPL` fits Cox proportional hazards models by **maximum penalised
likelihood** (MPL).  Unlike partial likelihood, MPL estimates the baseline
hazard $h_0$ jointly with the regression coefficients $\boldsymbol{\beta}$, by
expanding it in a basis of non-negative functions

$$h_0(t) = \sum_{u=1}^{m} \theta_u \psi_u(t), \qquad \boldsymbol{\theta} \geq \mathbf{0},$$

and penalising the roughness of the result.  Because $h_0$ is estimated rather
than profiled out, fits yield *absolute* hazard, cumulative hazard and survival
estimates, and the method extends naturally to censoring schemes the partial
likelihood cannot handle.

Supported data:

- right censoring, including left truncation (delayed entry)
- left censoring
- interval censoring
- any mixture of the above, together with exact event times

## Installation

From CRAN:

```r
install.packages("survivalMPL")
```

The development version:

```r
# install.packages("remotes")
remotes::install_github("survivalMPL/survivalMPL")
```

Then attach it alongside `survival`, which provides the `Surv()` response
constructor:

```{r load}
library(survivalMPL)
library(survival)
```

## The main function

`coxph_mpl()` is the entry point.  Its interface follows `survival::coxph()`:

```r
coxph_mpl(formula, data, subset, na.action, entry, control, ...)
```

| Argument | Purpose |
|---|---|
| `formula` | A model formula whose response is a `Surv()` object |
| `data` | Data frame containing the variables in `formula` |
| `entry` | Optional left-truncation (delayed entry) times |
| `control` | A list from `coxph_mpl.control()` |

The response tells `coxph_mpl()` which censoring scheme applies.  Right-censored
data can use the familiar two-argument form; everything else uses
`type = "interval2"`, where `NA` marks an unobserved endpoint:

| Data | Response |
|---|---|
| Right censored | `Surv(time, status)` |
| Left censored | `Surv(NA, t_R, type = "interval2")` |
| Interval censored | `Surv(t_L, t_R, type = "interval2")` |
| Exact event | `Surv(t, t, type = "interval2")` |

### Tuning the fit

`coxph_mpl.control()` collects everything governing the baseline hazard and the
optimiser.  The arguments most often changed are:

| Argument | Meaning |
|---|---|
| `basis` | Basis for $h_0$: `"uniform"`, `"gaussian"`, `"msplines"`, `"epanechikov"` |
| `n.knots` | Length-2 vector: numbers of quantile and equally spaced internal knots |
| `smooth` | Smoothing parameter $\lambda$; `NULL` estimates it by REML |
| `n.obs` | Number of exact events, used to set default knot counts |
| `max.iter` | Iteration limits for the three optimisation stages |

The registered bases are:

```{r bases}
list_bases()
```

### Methods

A fitted `coxph_mpl` object supports the usual generics:

| Call | Returns |
|---|---|
| `summary(fit)` | Coefficients, standard errors, and $\boldsymbol{\theta}$ |
| `coef(fit)` | Regression coefficients |
| `predict(fit, type = "risk")` | Estimated hazard $\hat h(t)$ |
| `predict(fit, type = "survival")` | Estimated survival $\hat S(t)$ |
| `plot(fit)` | Baseline hazard and survival with confidence bands |
| `residuals(fit)` | Martingale residuals |

## Where to go next

The tutorials work through one censoring scheme at a time — **right censoring**,
**left censoring**, and **interval censoring** — each with a complete worked
example.  Two further articles cover the modelling choices: **basis functions
for the baseline hazard**, and a side-by-side comparison of **`coxph` and
`coxph_mpl`** on the same data.

See `?coxph_mpl` and `?coxph_mpl.control` for the full argument lists.
