---
title: "Getting Started with survivalMPL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with survivalMPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

## Overview

`survivalMPL` fits Cox proportional hazards models with right, left, and interval censoring using maximum penalised likelihood (MPL). The baseline hazard is estimated with a smooth basis (step function, Gaussian, M-splines, or Epanechnikov) and non-negativity constraints are enforced during optimisation.

Below are two small examples: a right-censored fit on the `survival::lung` data and an interval-censored fit on the bundled `bcos2` data. To keep the vignette fast to build, we use modest iteration limits in `coxph_mpl.control()`.

## Right-censored example (`lung`)

```{r lung-fit}
library(survivalMPL)
library(survival)

lung <- na.omit(lung[, c("time", "status", "age", "sex", "ph.karno", "wt.loss")])

fit_lung <- coxph_mpl(
  Surv(time, status == 2) ~ age + sex + ph.karno + wt.loss,
  data = lung,
  control = coxph_mpl.control(
    n.obs = sum(lung$status == 2),
    max.iter = c(40, 2000, 4000),
    smooth = 0
  )
)

summary(fit_lung)
```

You can inspect the baseline hazard components or predict survival curves:

```{r lung-predict}
pred_lung <- predict(fit_lung, type = "survival")
head(pred_lung)
```

## Interval-censored example (`bcos2`)

```{r interval-fit}
data(bcos2)

fit_bcos <- coxph_mpl(
  Surv(left, right, type = "interval2") ~ treatment,
  data = bcos2,
  control = coxph_mpl.control(
    basis = "msplines",
    n.obs = nrow(bcos2),
    max.iter = c(40, 2000, 4000),
    smooth = 0
  )
)

summary(fit_bcos)
```

Predicted survival for the two treatment groups:

```{r interval-predict}
pred_bcos <- predict(fit_bcos, type = "survival", i = 1:2)
head(pred_bcos)
```

For richer diagnostics, see `plot(fit_bcos)`, `plot(predict(fit_bcos))`, and `plot(residuals(fit_lung))`.
