---
title: "Interval Censoring with survivalMPL"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Interval Censoring with survivalMPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5
)
set.seed(1)
```

## Overview

Survival times are *interval censored* when the event is known to have occurred
in an interval $(t_L, t_R]$ rather than at an exact time.  This is the normal
state of affairs in clinical studies where patients are examined only
periodically: a recurrence found at a follow-up visit happened at some point
since the previous visit, and nothing narrower is known.

The likelihood contribution is a difference of survivor functions,

$$P(t_L < T \leq t_R) = S(t_L) - S(t_R),$$

which requires $H_0$ at both endpoints and so cannot be written as a partial
likelihood.  `survivalMPL` handles *partly* interval-censored data: a dataset
may mix exact event times with left-, right- and interval-censored
observations, all in one fit.  The censoring type is communicated through
`survival::Surv(..., type = "interval2")`, which infers it from the pair of
endpoints:

| Data | Response | Status code |
|---|---|---|
| Exact event at $t$ | `Surv(t, t, type = "interval2")` | 1 |
| Interval censored on $(t_L, t_R]$ | `Surv(t_L, t_R, type = "interval2")` | 3 |
| Left censored at $t_R$ | `Surv(NA, t_R, type = "interval2")` | 2 |
| Right censored at $t_L$ | `Surv(t_L, NA, type = "interval2")` | 0 |

`melanoma` stores an unknown lower endpoint as `0` rather than `NA`, and an
unknown upper endpoint as `Inf` rather than `NA`.  `Inf` is read as right
censoring exactly like `NA`; `0` is read as an interval starting at zero, which
is the same statement as left censoring since $S(0) = 1$ makes
$S(0) - S(t_R)$ and $1 - S(t_R)$ identical.  The likelihood is unaffected; only
the status code `Surv()` reports differs.

This tutorial follows the pseudo-melanoma example of @MaWebbHudson2024
(Section 2.11).

---

## The `melanoma` data

`melanoma` is a pseudo dataset of time to first local recurrence for $n = 300$
melanoma patients, simulated from the design in @MaWebbHudson2024 with a
Weibull baseline hazard $h_0(t) = t^{-1/2}$.  Recurrence times are interval
censored because they are detected at clinic visits.  Covariates are the
melanoma **location** at first diagnosis (reference: head and neck), Breslow
**thickness** (reference: $< 1$ mm), **gender** (reference: male), and centred
**age** in decades.

```{r data}
library(survivalMPL)
library(survival)

data(melanoma)

str(melanoma)
```

Two thirds of the subjects have an exactly observed recurrence time; the
remaining third is censored, mostly by interval censoring:

```{r composition}
with(melanoma, c(
  exact    = sum(t_L == t_R & is.finite(t_R)),
  interval = sum(t_L > 0 & t_L < t_R & is.finite(t_R)),
  right    = sum(is.infinite(t_R)),
  left     = sum(t_L == 0 & is.finite(t_R))
))

table(status = Surv(melanoma$t_L, melanoma$t_R, type = "interval2")[, 3])
```

The status table counts 77 interval-censored observations rather than 58,
because the 19 left-censored rows are stored as intervals $(0, t_R]$ and
`Surv()` reports them as such.

The recurrence times are strongly right skewed: 95% of the observed endpoints
fall below one year, but a few reach four.

```{r skew}
endpoints <- c(melanoma$t_L[melanoma$t_L > 0],
               melanoma$t_R[is.finite(melanoma$t_R)])
round(quantile(endpoints, c(0.5, 0.9, 0.95, 1)), 3)
```

---

## Model fit

The call below is the one used in the book: cubic M-splines (`basis = "m"`) for
the baseline hazard, seven quantile knots, a convergence tolerance of $10^{-5}$
and generous iteration caps.  The smoothing value is not set, which leaves it to
the automatic smoothing parameter selection the package performs by default -
the marginal likelihood method of @MaWebbHudson2024 (Section 2.10), which they
single out as the package's default behaviour.

```{r fit}
mela_fit <- coxph_mpl(
  formula = Surv(t_L, t_R, type = "interval2") ~
    Arm + Leg + Trunk + mm1to2 + mm2to4 + mm4plus + Female + Age_centred,
  data = melanoma,
  basis = "m", tol = 1e-05, n.knots = c(7, 0),
  max.iter = c(1000, 5000, 1e+05)
)

summary(mela_fit)
```

The summary reports the fit on the log-hazard-ratio scale, together with the
$m = 10$ baseline hazard coefficients $\hat{\boldsymbol{\theta}}$ and the
selected smoothing value.  Convergence is reached well inside the iteration
caps.

### Hazard ratios

The book presents the same fit as hazard ratios with confidence intervals and
tests (its Table 2.6), which is the form a clinical reader expects:

```{r hazard-ratios}
b  <- coef(mela_fit)
se <- mela_fit$se$Beta$M2QM2
z  <- b / se

hr_tab <- data.frame(
  Covariate      = c("Arm", "Leg", "Trunk", "1 to 2 mm", "2 to 4 mm",
                     "4 mm and more", "Female", "Age (per 10 years)"),
  `HR estimate`  = round(exp(b), 3),
  `95% CI lower` = round(exp(b - qnorm(0.975) * se), 3),
  `95% CI upper` = round(exp(b + qnorm(0.975) * se), 3),
  `p-value`      = format.pval(2 * pnorm(-abs(z)), digits = 3, eps = 1e-4),
  check.names    = FALSE,
  row.names      = NULL
)

knitr::kable(hr_tab, align = "lrrrr")
```

Read against the head-and-neck reference, a melanoma first diagnosed on the arm
carries a significantly lower risk of recurrence, while the leg and trunk
categories are not distinguishable from the reference.  Breslow thickness is the
dominant risk factor: relative to tumours under 1 mm, the hazard is about 1.9
times higher at 2-4 mm and 3.5 times higher above 4 mm, both with $p < 0.001$.
A 10-year increase in age raises the recurrence hazard by about 10%, marginally
significant here.  Gender points the same way as in the book - women at lower
risk - but does not reach significance in this sample.

The book's Table 2.6 reports similar hazard ratios from its own simulated
sample, in which 37% of the recurrence times are exactly observed.  The bundled
`melanoma` is generated with two thirds exact events, so the point estimates are
comparable but the confidence intervals differ, and the trunk and gender effects
that are borderline in the book do not reach significance here.  The
coefficients used to generate the data are documented in `?melanoma`.

## Baseline hazard and survival

`plot()` on the fitted object gives the basis functions, the baseline hazard,
the cumulative baseline hazard and the baseline survival - the book's Figure 2.5
is the second of these:

```{r plot-all, fig.width=8, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(mela_fit, ask = FALSE, cex.main = 0.8)
```

At baseline covariate values the estimated hazard of recurrence falls steeply
and monotonically away from the origin and then flattens close to a constant,
which is the behaviour the true $h_0(t) = t^{-1/2}$ has.  The slight lift at the
extreme right, past four years, rests on a handful of observations and is not
supported by the confidence band.

Because the recurrence times are so skewed, everything of interest is squeezed
into the left-hand tenth of the panel.  `xlim` narrows the displayed range;
the fit itself, and the knots, are unchanged:

```{r plot-hazard, fig.width=6.5, fig.height=4.5}
plot(mela_fit, which = 2, ask = FALSE, cex.main = 0.9, xlim = c(0, 1.2))
```

Beyond the book's example, `predict()` turns the same fit into absolute survival
for a named covariate profile.  Contrasting the thinnest and thickest tumours
shows the size of the thickness effect on that scale:

```{r plot-predict, fig.width=6.5, fig.height=4.5}
i_thin  <- which(melanoma$mm1to2 == 0 & melanoma$mm2to4 == 0 &
                 melanoma$mm4plus == 0)[1]
i_thick <- which(melanoma$mm4plus == 1)[1]

p_thin  <- predict(mela_fit, type = "survival", i = i_thin)
p_thick <- predict(mela_fit, type = "survival", i = i_thick)

par(mar = c(4, 4.2, 3, 1))
plot(p_thin$time, p_thin$survival, type = "l", col = "steelblue", lwd = 2.4,
     xlim = c(0, 1.2), ylim = c(0, 1), las = 1,
     xlab = "Years since diagnosis", ylab = expression(hat(S)(t)),
     main = "Predicted recurrence-free survival", cex.main = 0.95)
lines(p_thick$time, p_thick$survival, col = "tomato", lwd = 2.4)
legend("topright", legend = c("Breslow thickness < 1 mm", "Breslow thickness > 4 mm"),
       col = c("steelblue", "tomato"), lwd = 2.4, bty = "n", cex = 0.85)
```

---

## Next steps

- **Left Censoring** covers the boundary case of interval censoring, where the
  lower endpoint is unknown, on a dataset with no interval-censored
  observations at all.
- **Right Censoring** covers the standard scheme and the `Surv(time, status)`
  response.
- **Control Parameters** explains `basis`, `n.knots`, `tol` and `max.iter`, the
  settings the fit above changes from their defaults.
- **Basis Functions for the Baseline Hazard** explains the four choices for
  $\psi_u$, including the `"msplines"` basis used above.

## References
