---
title: "Left Censoring with survivalMPL"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Left Censoring with survivalMPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5
)
set.seed(1)
```

## Overview

A survival time is *left censored* when the event is known to have happened
already by the time the subject is first examined, but not when.  All that is
recorded is an upper bound $t_R$: the event occurred somewhere in $(0, t_R]$.
This arises whenever the outcome is detected by inspection rather than observed
as it happens: a tumour already present at the first scan, seroconversion
already complete at enrolment, a milestone already reached at the first
assessment.

The likelihood contribution is the distribution function rather than the survivor
function,

$$P(T \leq t_R) = 1 - S(t_R) = 1 - \exp\bigl\{-H_0(t_R)e^{\mathbf{x}^T\boldsymbol{\beta}}\bigr\},$$

which the partial likelihood cannot accommodate.  Because `coxph_mpl()` estimates
$H_0$ explicitly, left-censored observations are handled directly, and may be
mixed freely with exact events and right- and interval-censored observations in
the same fit.

## Encoding left-censored observations

Left censoring uses the `type = "interval2"` response, with `NA` in place of the
unknown lower endpoint:

```{r encoding}
library(survivalMPL)
library(survival)

Surv(c(NA, 2, 3, 5), c(7, 4, NA, 5), type = "interval2")
```

The four observations above are, in order: left censored at 7 (`7-`), interval
censored on $(2, 4]$, right censored at 3 (`3+`), and an exact event at 5.
`Surv()` records the distinction in its status code, and `coxph_mpl()` dispatches
on that code:

```{r status-codes}
s <- Surv(c(NA, 2, 3, 5), c(7, 4, NA, 5), type = "interval2")
data.frame(status = s[, 3],
           meaning = c("left censored", "interval censored",
                       "right censored", "exact event"))
```

Note that `NA` here is *information*, not missingness: these rows are kept, not
dropped by `na.action`.

---

## Example: pseudo-melanoma data

The bundled `melanoma` data (time to first local recurrence, $n = 300$,
simulated) is partly interval censored and contains all four observation types.
Recurrences already present at the first assessment are recorded with a lower
endpoint of zero:

```{r melanoma-composition}
data(melanoma)

with(melanoma, c(
  exact    = sum(t_L == t_R & is.finite(t_R)),
  left     = sum(t_L == 0 & is.finite(t_R)),
  interval = sum(t_L > 0 & t_L < t_R & is.finite(t_R)),
  right    = sum(is.infinite(t_R))
))
```

So 132 of the 300 subjects (44%) are left censored.

### Two equivalent encodings

Recording a left-censored observation as the interval $(0, t_R]$ and recording it
as left censored at $t_R$ describe the same event, since $S(0) = 1$ makes
$S(0) - S(t_R)$ and $1 - S(t_R)$ identical.  The data ship in the first form; the
second is obtained by replacing the zero lower endpoints with `NA`:

```{r reencode}
mel <- melanoma
mel$t_L2 <- ifelse(melanoma$t_L == 0, NA, melanoma$t_L)
mel$t_R2 <- ifelse(is.finite(melanoma$t_R), melanoma$t_R, NA)

rbind(
  `as interval (0, t_R]` = table(factor(Surv(melanoma$t_L, melanoma$t_R,
                                            type = "interval2")[, 3],
                                        levels = 0:3)),
  `as left censored`     = table(factor(Surv(mel$t_L2, mel$t_R2,
                                            type = "interval2")[, 3],
                                        levels = 0:3))
)
```

The 132 observations move from status `3` (interval) to status `2` (left), and
fitting either version gives the same answer:

```{r fit-both}
ctrl <- coxph_mpl.control(
  basis    = "msplines",
  n.obs    = sum(melanoma$t_L == melanoma$t_R & is.finite(melanoma$t_R)),
  smooth   = 0,
  max.iter = c(40, 2e4, 5e4)
)

fit_interval <- coxph_mpl(
  Surv(t_L, t_R, type = "interval2") ~
    Arm + Leg + Trunk + mm1to2 + mm2to4 + mm4plus + Female + Age_centred,
  data = melanoma, control = ctrl)

fit_left <- coxph_mpl(
  Surv(t_L2, t_R2, type = "interval2") ~
    Arm + Leg + Trunk + mm1to2 + mm2to4 + mm4plus + Female + Age_centred,
  data = mel, control = ctrl)

round(cbind(`interval (0, t_R]` = coef(fit_interval),
            `left censored`     = coef(fit_left)), 3)
```

The two columns agree to about 0.02.  They are not bit-identical because the knot
positions are quantiles of the observed endpoint times, and the two encodings
present slightly different sets of endpoints, not because the likelihoods
differ.

### Inspecting the fit

```{r summary-left}
summary(fit_left)
```

The recurrence times are strongly right-skewed: 95% of the observed endpoints
fall below 1.1 years, but a few reach 4.  Over the full range every feature of
the fit is squeezed into the left-hand tenth of the panel, so we narrow the
displayed range with `xlim`.  This affects only what is drawn; the fit itself,
and the knots, are unchanged.

```{r plot-left, fig.width=8, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_left, ask = FALSE, cex.main = 0.8, xlim = c(0, 1.2))
```

The M-spline basis places most of its knots where the events are, so the first
panel is dense near zero.  The wide confidence band on $S_0(t)$ reflects how
little information the left-censored observations carry individually: each one
says only that the event happened at some point before $t_R$.

---

## Example: `bcos2`

The breast cosmesis data are mostly interval censored, but include a handful of
genuinely left-censored subjects, already coded with `NA`:

```{r bcos2}
data(bcos2)

table(status = Surv(bcos2$left, bcos2$right, type = "interval2")[, 3])
head(bcos2[is.na(bcos2$left), ])
```

Status `2` marks the left-censored rows; no special handling is needed:

```{r bcos2-fit}
fit_bcos <- coxph_mpl(
  Surv(left, right, type = "interval2") ~ treatment,
  data    = bcos2,
  control = coxph_mpl.control(
    basis    = "msplines",
    n.obs    = nrow(bcos2),
    max.iter = c(40, 2000, 4000),
    smooth   = 0
  )
)

summary(fit_bcos)
```

The same `plot()` method applies.  Here we show the basis functions and the
estimated baseline survival:

```{r bcos2-plot, fig.width=8, fig.height=3.8}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(fit_bcos, ask = FALSE, which = c(1, 4), cex.main = 0.85)
```

The hazard and cumulative hazard panels (`which = 2:3`) are omitted for this fit
rather than shown: with no exact event times and very little information in the
last knot interval, the delta-method standard error of the final
$\hat\theta_u$ is enormous, and the resulting confidence band dominates the
vertical scale to the point where the estimate itself is invisible.  The
survival panel is unaffected because it is bounded in $[0, 1]$.

---

## Next steps

- **Interval Censoring** covers the general partly interval-censored case, of
  which left censoring is the boundary instance.
- **Right Censoring** covers the standard scheme and left truncation.
- **Basis Functions for the Baseline Hazard** explains the choice of $\psi_u$;
  because $H_0(t_R)$ enters the left-censored likelihood directly, that choice
  matters more here than for right-censored data alone.
