---
title: "Left Truncation with survivalMPL"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Left Truncation with survivalMPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5
)
```

## Overview

A survival time is *left truncated* when any event occurring before some time
$a_i$ could never have been observed at all.  The truncation time is also called
a **delayed entry** time, because it usually arises from a condition on entering
the study: a subject who fails before meeting that condition never appears in the
data.

Truncation is not censoring, and the difference is fundamental.  Censoring means
a survival time is only *partially* observed: the subject is in the data, and
contributes the information that the event had not yet happened.  Truncation
means an event time that could have been observed becomes *entirely*
unobservable when it falls before $a_i$: the subject is absent from the data
altogether.  Treating a truncation time as time zero for follow-up, and so
ignoring the truncation, leads to biased parameter estimates
[@BhaskaranLiquetMa].

Each subject therefore contributes $(a_i, y_i, \delta_i, \mathbf{x}_i)$: the
truncation time, the observed event or censoring time, the event indicator, and
the covariates, with $a_i$ strictly less than $y_i$.

## Encoding left-truncated observations

Truncation times are not part of the `Surv()` response.  Pass them through the
`entry` argument instead:

```r
coxph_mpl(Surv(time, status) ~ x, data = df, entry = start)
```

Three things to know:

- **`entry` requires `basis = "uniform"`**, which is the default, so nothing
  else has to be set. Left truncation is implemented by
  differencing the cumulative basis, $H_0(a_i, y_i) = H_0(y_i) - H_0(a_i)$, which
  the package supports for the piecewise-constant basis.  This is not a
  limitation of the method so much as its natural setting: with indicator basis
  functions $\psi_u(t) = I(t \in B_u)$ over bins $B_u = (w_u, w_{u+1}]$, each
  $\hat\theta_u$ has a closed form given $\boldsymbol{\beta}$, which is what
  makes the profile likelihood of @BhaskaranLiquetMa possible.  Asking for any
  other basis together with `entry` gets you a uniform fit and a warning saying
  so, rather than a silently wrong fit or a refusal.
- **Entry times must strictly precede the event or censoring time.** A violation
  is an error, not a warning.
- **Truncated and untruncated subjects may be mixed** in one call. Use an entry
  time of zero, or any value below the first event time, for subjects observed
  from the origin.

## Example: atomic bomb survivors

The bundled `hiroshima` data are left truncated by construction.  Follow-up in
the Life Span Study begins on 1 October 1950, several years after the 1945
exposure, so a subject enters the risk set at the age they had reached by that
date, and anyone who died in between is absent from the cohort.  `entry` is that
attained age, and `time` is attained age at death or at end of follow-up.

```{r data}
library(survivalMPL)
library(survival)

data(hiroshima)

c(subjects = nrow(hiroshima), deaths = sum(hiroshima$status))
round(range(hiroshima$entry), 1)
round(range(hiroshima$time), 1)
```

Entry ages run from 5 to 94 years, so the truncation is substantial rather than
incidental: at any attained age, the subjects at risk are only those who had
already reached that age by late 1950.

```{r fit-entry}
fit_entry <- coxph_mpl(
  Surv(time, status) ~ dose + sex + city,
  data  = hiroshima,
  entry = entry
)

summary(fit_entry)
```

The estimated baseline hazard is a hazard by *attained age*, and because the
risk set at each age contains only subjects who had already reached that age by
late 1950, it is the truncation-corrected one:

```{r plot-entry-hazard, fig.width=6.5, fig.height=4.5}
plot(fit_entry, which = 2, ask = FALSE, cex.main = 0.9)
```

Requesting a different basis alongside `entry` does not fail; the fit falls back
to `"uniform"` and says so:

```{r basis-fallback}
fallback <- coxph_mpl(
  Surv(time, status) ~ dose + sex + city,
  data  = hiroshima,
  entry = entry,
  basis = "msplines"
)

fallback$control$basis
```

## Limitations

`predict.coxph_mpl()` and `residuals.coxph_mpl()` do not yet account for
`entry`: they compute cumulative hazard and survival from time zero rather than
from each subject's entry time.  Coefficients, baseline hazard estimates and
their standard errors are unaffected.  See `?coxph_mpl` for the current status.

## References
