---
title: "Left Censoring with survivalMPL"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Left Censoring with survivalMPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5
)
set.seed(1)
```

## Overview

A survival time is *left censored* when the event is known to have happened
already by the time the subject is first examined, but not when.  All that is
recorded is an upper bound $t_R$: the event occurred somewhere in $(0, t_R]$.
This arises whenever the outcome is detected by inspection rather than observed
as it happens - a tumour already present at the first scan, seroconversion
already complete at enrolment, a milestone already reached at the first
assessment.

The likelihood contribution is the distribution function rather than the
survivor function,

$$P(T \leq t_R) = 1 - S(t_R) = 1 - \exp\bigl\{-H_0(t_R)e^{\mathbf{x}^T\boldsymbol{\beta}}\bigr\},$$

which the partial likelihood cannot accommodate: it needs a risk set at a known
event time, and a left-censored subject supplies neither.  Because
`coxph_mpl()` estimates $H_0$ explicitly, left-censored observations enter the
likelihood directly, and may be mixed freely with exact events and right- and
interval-censored observations in the same fit.

## Encoding left-censored observations

Left censoring uses the `type = "interval2"` response, with `NA` in place of the
unknown lower endpoint:

```{r encoding}
library(survivalMPL)
library(survival)

Surv(c(NA, 2, 3, 5), c(7, 4, NA, 5), type = "interval2")
```

The four observations above are, in order: left censored at 7 (`7-`), interval
censored on $(2, 4]$, right censored at 3 (`3+`), and an exact event at 5.
`Surv()` records the distinction in its status code, and `coxph_mpl()` dispatches
on that code:

```{r status-codes}
s <- Surv(c(NA, 2, 3, 5), c(7, 4, NA, 5), type = "interval2")
data.frame(status = s[, 3],
           meaning = c("left censored", "interval censored",
                       "right censored", "exact event"))
```

Note that `NA` here is *information*, not missingness: these rows are kept, not
dropped by `na.action`.

Recording a left-censored observation as `NA` and recording it as the interval
$(0, t_R]$ describe the same event, since $S(0) = 1$ makes $1 - S(t_R)$ and
$S(0) - S(t_R)$ identical.  Both encodings are accepted; the `NA` form is the
one that states the intent.

---

## Example: HIV seroconversion

The bundled `hiv` data are entirely synthetic - there is no real cohort behind
them, and the coefficients are chosen for illustration rather than estimated
from any study.  The design imitates a cohort of people who inject drugs,
followed for HIV seroconversion, with time measured in years from the start of
injecting drug use.  The observation scheme is what makes it useful here: each
subject has a first HIV test some time after that origin, and

- a subject already seropositive at the first test seroconverted somewhere in
  $(0, t_R]$ - **left censored**;
- a subject negative at the first test is retested frequently, so a later
  seroconversion is recorded as an **exact event**;
- a subject still negative when follow-up ends is **right censored**.

No subject is interval censored, so this is left censoring on its own rather
than as a special case of something more general.

```{r hiv-composition}
data(hiv)

str(hiv)

table(status = Surv(hiv$t_L, hiv$t_R, type = "interval2")[, 3])
```

Status `2` marks the left-censored rows (96 subjects, just under a third of the
sample), status `1` the exact events and status `0` the right-censored rows.
The left-censored subjects are the ones with no lower bound at all:

```{r hiv-head}
head(hiv[is.na(hiv$t_L), ])
```

### Model fit

$H_0$ is evaluated at $t_R$ for every left-censored subject, so the shape of the
baseline matters more here than it does for right-censored data, where the
likelihood only ever looks at the hazard at the event times.  That argues for a
smooth basis: `basis = "msplines"`.  The knot count is set with `n.knots`:
a third of this sample contributes only an upper bound, which is not enough
information to support the default knot sequence, so six quantile knots are
used instead.  Everything else, including the automatic selection of the
smoothing value, is left at its default - see **Control Parameters**.

```{r hiv-fit}
fit_hiv <- coxph_mpl(
  Surv(t_L, t_R, type = "interval2") ~ Sharing + Prison + Female + Age_centred,
  data    = hiv,
  basis   = "msplines",
  n.knots = c(5, 0)
)

summary(fit_hiv)
```

The data were generated with true coefficients $0.90$ for `Sharing`, $0.45$ for
`Prison`, $-0.25$ for `Female` and $-0.15$ for `Age_centred`, so the fit can be
checked against the truth:

```{r hiv-truth}
truth <- c(Sharing = 0.90, Prison = 0.45, Female = -0.25, Age_centred = -0.15)

round(cbind(estimate = coef(fit_hiv),
            SE       = fit_hiv$se$Beta$M2HM2,
            true     = truth), 3)
```

### Baseline hazard and survival

```{r hiv-plot, fig.width=8, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_hiv, ask = FALSE, cex.main = 0.8)
```

The four panels are the M-spline basis functions, then the estimated baseline
hazard, cumulative hazard and survival.  The hazard panel is the one to read
first: the true baseline is Weibull with shape 1.5, an increasing hazard, and
the estimate reproduces that rise.  The confidence band is wide, which is
honest - a left-censored subject contributes only the statement that the event
happened at some point before $t_R$, so a third of this sample carries very
little information about *when*.

```{r hiv-hazard, fig.width=6.5, fig.height=4.5}
plot(fit_hiv, which = 2, ask = FALSE, cex.main = 0.9)
```

### Predicted survival

Because $H_0$ is estimated, survival can be predicted for a named covariate
profile rather than only compared between profiles.  Here are two subjects who
differ in the strongest covariate:

```{r hiv-predict, fig.width=6.5, fig.height=4.5}
i_share    <- which(hiv$Sharing == 1)[1]
i_no_share <- which(hiv$Sharing == 0)[1]

p1 <- predict(fit_hiv, type = "survival", i = i_share)
p0 <- predict(fit_hiv, type = "survival", i = i_no_share)

par(mar = c(4, 4.2, 3, 1))
plot(p0$time, p0$survival, type = "l", col = "steelblue", lwd = 2.4,
     ylim = c(0, 1), xlab = "Years since first injecting drug use",
     ylab = expression(hat(S)(t)), las = 1,
     main = "Predicted survival free of seroconversion", cex.main = 0.95)
lines(p1$time, p1$survival, col = "tomato", lwd = 2.4)
legend("bottomleft", legend = c("no equipment sharing", "equipment sharing"),
       col = c("steelblue", "tomato"), lwd = 2.4, bty = "n", cex = 0.85)
```

---

## Next steps

- **Interval Censoring** covers the general partly interval-censored case, of
  which left censoring is the boundary instance, and uses a dataset containing
  all four observation types at once.
- **Right Censoring** covers the standard scheme and the `Surv(time, status)`
  response.
- **Control Parameters** explains `n.knots`, `basis` and `smooth`, the settings
  that decide how much structure the baseline hazard is allowed to have.
- **Basis Functions for the Baseline Hazard** explains the choice of $\psi_u$;
  because $H_0(t_R)$ enters the left-censored likelihood directly, that choice
  matters more here than for right-censored data alone.

## References
