---
title: "Right Censoring with survivalMPL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Right Censoring with survivalMPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center",
                      fig.width = 7, fig.height = 4.5)
set.seed(1)
```

## Overview

A survival time is *right censored* when follow-up ends before the event occurs:
all we know is that the event time exceeds the last observation time.  This is
the most common censoring scheme, and the one `survival::coxph()` handles through
the partial likelihood.

`coxph_mpl()` fits the same proportional hazards model, but estimates the
baseline hazard $h_0$ jointly with $\boldsymbol{\beta}$ instead of profiling it
out.  The practical consequence is that a fit gives absolute hazard and survival
estimates directly, with standard errors, rather than requiring a separate
Breslow-type step afterwards.

Right-censored data use the ordinary two-argument response, exactly as in
`coxph()`:

```r
Surv(time, status)
```

where `status` is `1`/`TRUE` for an observed event and `0`/`FALSE` for a censored
observation.

---

## Example — `survival::lung`

The `lung` data from the **survival** package records survival in patients with
advanced lung cancer.  `status` is coded `2` for death and `1` for censored, so
the event indicator is `status == 2`.

```{r data}
library(survivalMPL)
library(survival)

lung <- na.omit(lung[, c("time", "status", "age", "sex", "ph.karno", "wt.loss")])

c(n = nrow(lung), events = sum(lung$status == 2),
  censored = sum(lung$status == 1))
```

### Model fit

`n.obs` tells `coxph_mpl.control()` how many exact events there are, which sets
the default number of knots for the baseline hazard.  Here we fix the smoothing
parameter at `smooth = 0` (no penalty) to keep the example fast; leaving
`smooth = NULL` estimates it by REML instead.

```{r fit}
fit_lung <- coxph_mpl(
  Surv(time, status == 2) ~ age + sex + ph.karno + wt.loss,
  data = lung,
  control = coxph_mpl.control(
    n.obs    = sum(lung$status == 2),
    max.iter = c(40, 2000, 4000),
    smooth   = 0
  )
)

summary(fit_lung)
```

The coefficient table is read as in any Cox fit: `sex` is strongly protective
(coded 1 = male, 2 = female), and higher `ph.karno` — better performance status —
is associated with lower hazard.  The summary also reports the estimated
baseline hazard coefficients $\boldsymbol{\theta}$, which have no counterpart in
a partial likelihood fit.

### Baseline hazard and survival

Because $h_0$ is part of the model, it can be plotted with confidence bands
straight from the fitted object:

```{r plot-baseline, fig.height=4}
plot(fit_lung)
```

### Predicted survival

`predict()` returns absolute estimates with standard errors and pointwise
confidence limits.  With no `i` argument it evaluates at the mean covariate
vector:

```{r predict}
pred_lung <- predict(fit_lung, type = "survival")
head(pred_lung)
```

A single subject is selected with `i` (one row index at a time — only the first
element is used):

```{r predict-i}
pred_one <- predict(fit_lung, type = "survival", i = 1)
head(pred_one)
```

### Residuals

Martingale residuals are available for checking the fit:

```{r resid, fig.height=4}
plot(residuals(fit_lung))
```

---

## Left truncation (delayed entry)

Right-censored data are often also *left truncated*: subjects enter the risk set
only at some entry time, and anyone who fails before that time is never
observed.  Pass those times through `entry`:

```r
coxph_mpl(Surv(time, status) ~ x, data = df, entry = df$start,
          control = coxph_mpl.control(basis = "uniform", n.obs = sum(df$status)))
```

Entry times must strictly precede the observed event or censoring time, and left
truncation currently requires `basis = "uniform"`.  Truncated and untruncated
subjects may be mixed in the same call.  See `?coxph_mpl` for details.

---

## Next steps

- **Left Censoring** and **Interval Censoring** cover the other censoring
  schemes, which use the `type = "interval2"` response.
- **Basis Functions for the Baseline Hazard** explains the four choices for
  $\psi_u$ and how the knots are placed.
- **Comparing `coxph` and `coxph_mpl`** fits both to the same right-censored
  data and contrasts the results.
