---
title: "Right Censoring with survivalMPL"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Right Censoring with survivalMPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5
)
set.seed(1)
```

## Overview

A survival time is *right censored* when follow-up ends before the event occurs:
all we know is that the event time exceeds the last observation time.  This is
the most common censoring scheme, and the one `survival::coxph()` handles through
the partial likelihood.

`coxph_mpl()` fits the same proportional hazards model, but estimates the
baseline hazard $h_0$ jointly with $\boldsymbol{\beta}$ instead of profiling it
out.  The practical consequence is that a fit gives absolute hazard and survival
estimates directly, with standard errors, rather than requiring a separate
Breslow-type step afterwards.

## Encoding right-censored observations

Right-censored data use the ordinary two-argument response, exactly as in
`coxph()`:

```r
Surv(time, status)
```

where `status` is `1`/`TRUE` for an observed event and `0`/`FALSE` for a censored
observation.  This is the one scheme that does not need
`type = "interval2"`; internally `coxph_mpl()` converts it to that form anyway.

---

## Example: `survival::lung`

The `lung` data from the **survival** package records survival in patients with
advanced lung cancer.  `status` is coded `2` for death and `1` for censored, so
the event indicator is `status == 2`.

```{r data}
library(survivalMPL)
library(survival)

lung <- na.omit(lung[, c("time", "status", "age", "sex", "ph.karno", "wt.loss")])

c(n = nrow(lung), events = sum(lung$status == 2),
  censored = sum(lung$status == 1))
```

### Model fit

`n.obs` tells `coxph_mpl.control()` how many exact events there are, which sets
the default number of knots for the baseline hazard.  Here we fix the smoothing
parameter at `smooth = 0` (no penalty) to keep the example fast; leaving
`smooth = NULL` estimates it by REML instead.

```{r fit}
fit_lung <- coxph_mpl(
  Surv(time, status == 2) ~ age + sex + ph.karno + wt.loss,
  data = lung,
  control = coxph_mpl.control(
    n.obs    = sum(lung$status == 2),
    max.iter = c(40, 2000, 4000),
    smooth   = 0
  )
)

summary(fit_lung)
```

The coefficient table is read as in any Cox fit: `sex` is strongly protective
(coded 1 = male, 2 = female), and higher `ph.karno` (better performance status)
is associated with lower hazard.  The summary also reports the estimated
baseline hazard coefficients $\boldsymbol{\theta}$, which have no counterpart in
a partial likelihood fit.

### Baseline hazard and survival

Because $h_0$ is part of the model, it can be plotted with confidence bands
straight from the fitted object.  `plot()` produces four panels: the basis
functions, then the estimated baseline hazard, cumulative hazard and survival.

```{r plot-baseline, fig.width=8, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_lung, ask = FALSE, cex.main = 0.8)
```

In the first panel each of the $m$ basis functions is drawn in its own colour,
running through `terrain.colors()` from the earliest to the latest, so the
colour simply indexes $u$ in $\psi_u$ and carries no other meaning.  With the
default uniform basis each $\psi_u$ is an indicator on one knot interval, so
the panel reads as a row of adjacent boxes of height 1, one colour per
interval; the pale grey verticals are the knots $\boldsymbol{\alpha}$.  A basis
drawn dashed would indicate $\hat\theta_u$ driven to zero, meaning that
interval contributes nothing to the fitted hazard.

### Predicted survival

`predict()` returns absolute estimates with standard errors and pointwise
confidence limits.  With no `i` argument it evaluates at the mean covariate
vector:

```{r predict}
pred_lung <- predict(fit_lung, type = "survival")
head(pred_lung)
```

A single subject is selected with `i` (one row index at a time; only the first
element is used):

```{r predict-i}
pred_one <- predict(fit_lung, type = "survival", i = 1)
head(pred_one)
```

### Residuals

Martingale residuals are available for checking the fit:

```{r resid, fig.height=4}
plot(residuals(fit_lung))
```

---

## Next steps

- **Left Truncation** covers delayed entry, where subjects join the risk set
  only at some entry time.  Right-censored data are often also left truncated,
  and `coxph_mpl()` handles it through the `entry` argument.
- **Left Censoring** and **Interval Censoring** cover the other censoring
  schemes, which use the `type = "interval2"` response.
- **Basis Functions for the Baseline Hazard** explains the four choices for
  $\psi_u$ and how the knots are placed.
- **Comparing `coxph()` and `coxph_mpl()`** fits both to the same
  right-censored data and contrasts the results.
