---
title: "Right Censoring with survivalMPL"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Right Censoring with survivalMPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5
)
set.seed(1)
```

## Overview

A survival time is *right censored* when follow-up ends before the event occurs:
all we know is that the event time exceeds the last observation time.  This is
the most common censoring scheme, and the one `survival::coxph()` handles through
the partial likelihood.

`coxph_mpl()` fits the same proportional hazards model, but estimates the
baseline hazard $h_0$ jointly with $\boldsymbol{\beta}$ instead of profiling it
out.  The practical consequence is that a fit gives absolute hazard and survival
estimates directly, with standard errors, rather than requiring a separate
Breslow-type step afterwards.

## Encoding right-censored observations

Right-censored data use the ordinary two-argument response, exactly as in
`coxph()`:

```r
Surv(time, status)
```

where `status` is `1`/`TRUE` for an observed event and `0`/`FALSE` for a censored
observation.  This is the one scheme that does not need
`type = "interval2"`; internally `coxph_mpl()` converts it to that form anyway.

---

## Example: `survival::lung`

The `lung` data from the **survival** package records survival in patients with
advanced lung cancer.  `status` is coded `2` for death and `1` for censored, so
the event indicator is `status == 2`.

```{r data}
library(survivalMPL)
library(survival)

lung <- na.omit(lung[, c("time", "status", "age", "sex", "ph.karno", "wt.loss")])

c(n = nrow(lung), events = sum(lung$status == 2),
  censored = sum(lung$status == 1))
```

### Model fit

Little beyond the formula and the data is required.  The defaults use a uniform
(piecewise-constant) baseline hazard, take the number of knots from the number
of events, and select the smoothing value $\lambda$ automatically by the
marginal likelihood method.  The one change made here is the convergence
tolerance, set to the $10^{-5}$ used throughout @MaWebbHudson2024; at the
default $10^{-7}$ the outer iterations do not settle on this dataset - see
**Control Parameters**:

```{r fit}
fit_lung <- coxph_mpl(
  Surv(time, status == 2) ~ age + sex + ph.karno + wt.loss,
  data = lung,
  tol  = 1e-05
)

summary(fit_lung)
```

The coefficient table is read as in any Cox fit: `sex` is strongly protective
(coded 1 = male, 2 = female), and higher `ph.karno` (better performance status)
is associated with lower hazard.  The summary also reports the estimated
baseline hazard coefficients $\boldsymbol{\theta}$, which have no counterpart in
a partial likelihood fit.

Every one of those defaults can be overridden through `coxph_mpl.control()` -
the basis, the knots, the smoothing parameter, the iteration limits.  The
**Control Parameters** article covers them one at a time.

### Baseline hazard and survival

Because $h_0$ is part of the model, it can be plotted with confidence bands
straight from the fitted object.  `plot()` produces four panels: the basis
functions, then the estimated baseline hazard, cumulative hazard and survival.

```{r plot-baseline, fig.width=8, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_lung, ask = FALSE, cex.main = 0.8)
```

In the first panel each of the $m$ basis functions is drawn in its own colour,
running through `terrain.colors()` from the earliest to the latest, so the
colour simply indexes $u$ in $\psi_u$ and carries no other meaning.  With the
default uniform basis each $\psi_u$ is an indicator on one knot interval, so
the panel reads as a row of adjacent boxes of height 1, one colour per
interval; the pale grey verticals are the knots $\boldsymbol{\alpha}$.  A basis
drawn dashed would indicate $\hat\theta_u$ driven to zero, meaning that
interval contributes nothing to the fitted hazard.

The second panel is the estimated baseline hazard itself: a step function whose
height on knot interval $u$ is $\hat\theta_u$.  It is the one quantity a
partial likelihood fit cannot give you, so it is worth drawing on its own:

```{r plot-hazard-only, fig.width=6.5, fig.height=4.5}
plot(fit_lung, which = 2, ask = FALSE, cex.main = 0.9)
```

The hazard climbs steadily across follow-up - more than doubling between the
first knot interval and the last - and flattens only in the final intervals,
where few patients remain at risk and the confidence band widens accordingly.

### Predicted survival

`predict()` returns absolute estimates with standard errors and pointwise
confidence limits.  With no `i` argument it evaluates at the mean covariate
vector:

```{r predict}
pred_lung <- predict(fit_lung, type = "survival")
head(pred_lung)
```

A single subject is selected with `i` (one row index at a time; only the first
element is used):

```{r predict-i}
pred_one <- predict(fit_lung, type = "survival", i = 1)
head(pred_one)
```

---

## Residuals

`residuals()` returns both Cox-Snell and martingale residuals.  For a subject
observed exactly, the Cox-Snell residual is the fitted cumulative hazard that
subject accumulated,

$$r^{CS}_i = e^{\mathbf{x}_i^T\hat{\boldsymbol{\beta}}}\,\hat H_0(t_i),$$

with censored observations adjusted so that every residual estimates the same
quantity under the model.  The martingale residual is then
$\hat M_i = \delta_i - r^{CS}_i$: the difference between the number of events
actually observed for subject $i$ (0 or 1) and the number the model expects.

```{r resid, fig.width=8, fig.height=4}
res_lung <- residuals(fit_lung)
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(res_lung, which = 1:2, ask = FALSE)
```

Read them as follows.

- **Martingale residuals** are bounded above by 1 and unbounded below, so the
  cloud is asymmetric by construction: censored observations (red) are at or
  below zero, events (black) are pulled towards 1.  A residual close to 1 marks
  a subject who died far earlier than the model predicted, a large negative one
  a subject who survived far longer.  Here a handful of subjects sit below
  $-2$, one of them near $-6$: those are the long survivors the fitted model
  finds most surprising, and they are worth looking at individually before
  concluding anything about the model as a whole.
- **Cox-Snell residuals** behave like a censored sample from a unit exponential
  distribution when the model is correct, so most sit below 1 with a thinning
  right tail.  A mass of residuals far above 1 indicates the fitted cumulative
  hazard is wrong, usually through a mis-specified covariate effect rather than
  a mis-specified baseline.

The horizontal axis of both panels is the row number, which carries meaning
only if the rows arrived in a meaningless order.  In `lung` they did not: the
censored subjects are concentrated in the later rows (3 of the first 54 are
censored, against 36 of the last 53), which is why the red points drift to the
right of both panels.  That is a fact about the data file, not about the fit.

---

## Next steps

- **Control Parameters** covers `coxph_mpl.control()`: the basis, the knots,
  the smoothing value $\lambda$, the iteration limits and the convergence
  tolerance used above.
- **Left Truncation** covers delayed entry, where subjects join the risk set
  only at some entry time.  Right-censored data are often also left truncated,
  and `coxph_mpl()` handles it through the `entry` argument.
- **Left Censoring** and **Interval Censoring** cover the other censoring
  schemes, which use the `type = "interval2"` response.
- **Basis Functions for the Baseline Hazard** explains the four choices for
  $\psi_u$ and how the knots are placed.
- **Comparing `coxph()` and `coxph_mpl()`** fits both to the same
  right-censored data and contrasts the results.

## References
